---
title: 'Case Study: Bike-sharing'
output:
  html_document: default
  pdf_document: default
date: '2022-03-09'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Citations
From Google's Data Analysis Professional Certification

## Scenario
You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

## Business Question
How does a bike-share navigate speedy success?

## Other questions
- Is there a duration difference between members and casual riders?
- Is there a duration difference between regular versus motorized bike?
- Is there a distance difference between regular versus motorized bike?
- Is there a time of the week difference between members and casual riders?
- How has membership and casual riders changed over time?
- Is there a location difference between members and casual riders? Could this be used to market directly to casual?
- How does weather and temperature effect ridership of member versus casual riders?

## Import R libraries
```{r, echo=FALSE}
library(tidyverse)
library(lubridate)

```


## Import dataset

```{r echo=FALSE}
# Import Dataset
list_of_files <- list.files(path="tripdata",
                       recursive = T,
                       pattern ="\\.csv$",
                       full.names = T)

trip_data <- readr::read_csv(list_of_files, id = "file_name")
trip_data_t <- as_tibble(trip_data) # convert to tibble
```

## Summary Statistics of raw data

```{r echo=FALSE}
paste("Date range: ", trip_data_t$started_at[1], " to ", trip_data_t$started_at[nrow(trip_data_t)])
table(trip_data_t$member_casual)
```
## Subset the data, make calculations, and explore

- subset raw data so only the last 12 months are analyzed
- Calculate the ride duration
- Calculate the month, day of the week, and time of day of the trip

```{r echo=FALSE}
trip_data_previous_12m <- trip_data_t %>% 
                          filter(started_at >= as.Date("2021-03-01"))

# Calculate what day of the week the ride was on
trip_data_previous_12m$week_day <- wday(trip_data_previous_12m$started_at, label=TRUE)

# Calculate the month the ride was on
trip_data_previous_12m$month <- month(trip_data_previous_12m$started_at, label=TRUE)

### How long was the duration of bike rental?
trip_data_previous_12m$ride_duration <- as.numeric(difftime(trip_data_previous_12m$ended_at, trip_data_previous_12m$started_at))

glimpse(trip_data_previous_12m)

```
                          


```{r echo=FALSE}
# Clean data
# Remove duplicated observations and without geolocation data

trip_data_clean <- trip_data_previous_12m %>% 
  distinct(ride_id,  .keep_all = TRUE) %>%
  drop_na(end_lat, end_lng)

dropped_observations <- nrow(trip_data_previous_12m) - nrow(trip_data_clean)

# How many possible values for type of bike?
unique(trip_data_clean$rideable_type) # electric_bike, #docked_bike, #classic_bike



#Remove observations with negative duration & less than 5 seconds (10090 observations)
trip_data_clean_v2 <- trip_data_clean[(!trip_data_clean$ride_duration < 5), ]
dropped_less_1min <- nrow(trip_data_clean) - nrow(trip_data_clean_v2)

#Remove observations with duration greater than 12 hours (after 24 hours will be charged $1200 for the bike)
dropped_greater_12hour <- nrow(trip_data_clean_v2[trip_data_clean_v2$ride_duration > 12*60*60,])
trip_data_clean_v2 <- trip_data_clean_v2[!trip_data_clean_v2$ride_duration > 12*60*60,] #4248 removed

#Remove for more than 3 hours (19246 observations)
dropped_greater_3hour <- nrow(trip_data_clean_v2[trip_data_clean_v2$ride_duration > 3*60*60,])
trip_data_clean_v2 <- trip_data_clean_v2[!trip_data_clean_v2$ride_duration > 3*60*60,]

# Create subcategories of usertype and duration
trip_data_clean_v2$duration_factor <- ceiling(trip_data_clean_v2$ride_duration) < 30*60
trip_data_clean_v2 <- trip_data_clean_v2 %>% 
                      mutate(
                        duration_factor = ifelse(duration_factor == TRUE, "< 30", "< 180"))

```
## Clean and explore data

- No duplicated data
- Some entries have missing docking station data
   - Use GPS data for location analysis
   - Removed `r as.numeric(dropped_observations)` ride entries without geolocation data
- Removed `r as.numeric(dropped_less_1min)` ride entries with durations less than 1 minute
- Removed ride entries with durations greater than 3 hours
   - Removed `r as.numeric(dropped_greater_12hour)` trips with duration greater than 12 hours
   - Removed `r as.numeric(dropped_greater_3hour)` trips with duration greater than 3 hours

## Explore Cleaned data

`r table(trip_data_clean_v2$member_casual)`

### Median and Mean trip duration in minutes
`r trip_data_clean_v2 %>% group_by(member_casual) %>% summarize(Median = median(ride_duration)/60, Mean = mean(ride_duration)/60)`


```{r echo = FALSE}

plot_member <- ggplot(data = trip_data_clean_v2, aes(x = member_casual, fill=member_casual)) +
                  geom_bar(show.legend=FALSE) +
                  scale_y_continuous(labels = scales::comma) +
                  ggtitle("Total rides of members and casual users") +
                  theme(plot.title = element_text(hjust = 0.5)) +
                  ylab("Number of rides") + xlab("User type")

plot_member
```

### Frequency distribution of trip duration shows a right-skewed distribution

```{r echo=FALSE}
# Histogram of casual users and members trip duration
plot_duration <- ggplot(data = trip_data_clean_v2, 
                        aes(x = ceiling(ride_duration/60), fill=member_casual, shape=member_casual)) +
                 geom_bar(position="dodge") +
                 ggtitle("Distribution of ride duration of members and casual users") +
                 xlab("ride duration (min)") + ylab("count") +
                 theme(legend.position = c(.55, .75), 
                       legend.text = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5)) +
                labs(fill='User type') 
plot_duration

```

### Pearson's coefficient of skewness:


#### Boxplot indicates most rides are short duration

```{r echo = FALSE}
# Normal boxplot
boxplot_duration <- ggplot(data = trip_data_clean_v2, #head(trip_data_clean_v2, 1000), 
                        aes(x = member_casual, y = (ride_duration/60))) + 
                      geom_boxplot(alpha=0.3) + 
                      scale_fill_brewer(palette="BuPu") +
                      ggtitle("Boxplot of ride duration of members and casual users") +
                      theme(plot.title = element_text(hjust = 0.5)) +
                      ylab("ride duration (min)") + xlab("")

boxplot_duration
```

## Log transformation of data

### Log transforming removes the skewness of data

```{r echo = FALSE}
trip_data_clean_v2$ride_duration_log <- log(trip_data_clean_v2$ride_duration, 10)

theme_set(theme_classic())
plot_duration_log <- ggplot(data = trip_data_clean_v2, 
                        aes(x = ride_duration_log, fill=member_casual, shape=member_casual)) +
                 geom_density(alpha = 0.3) +
                 ggtitle("Probability distribution of log10 transformed ride duration") +
                 xlab("ride duration (min)") + ylab("count") +
                 theme(legend.position = c(.25, .65), 
                       legend.text = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5)) +
                labs(fill='User type') 

plot_duration_log




# Boxplot of log transformed data
boxplot_duration_log <- ggplot(data = trip_data_clean_v2, #head(trip_data_clean_v2, 1000), 
                        aes(x = member_casual, y = ride_duration_log, fill = member_casual)) + 
                        geom_boxplot(alpha=0.3) + 
                        ggtitle("Boxplot of ride duration of members and casual users") +
                        theme(legend.position = c(.85, .25), 
                              legend.text = element_text(face = "bold"),
                              plot.title = element_text(hjust = 0.5)) +
                        labs(fill='User type') +
                        ylab("log(ride duration)") + xlab("")

boxplot_duration_log

t.test(trip_data_clean_v2$ride_duration_log~trip_data_clean_v2$member_casual)
```



## Explore data by month, day-of-the-week, and hour-of-the-day

```{r echo=FALSE}
library(scales)

theme_set(theme_classic())




plot_yearly_use <- ggplot(data = trip_data_clean_v2, aes(x = year(started_at), fill=member_casual)) +
                   geom_bar(position="dodge") +
                   ggtitle("Last 12 month's yearly ridership of member and casual users") +
                   scale_y_continuous(labels = scales::comma) +
                        theme(legend.position = c(.85, .85), 
                              legend.text = element_text(face = "bold"),
                              plot.title = element_text(hjust = 0.5)) +
                        labs(fill='User type') +
                        ylab("log(ride duration)") + xlab("")

# Plot monthly use of casual versus members for last 12 months
# plot_monthly_use <- ggplot(data = trip_data_clean_v2, aes(x = as.Date(started_at, "%Y-%m-%d %h:%m:%s", tz="UTC"), fill=member_casual)) +
#                     geom_bar(position="dodge") +
#                     ggtitle("Last 12 month's monthly ridership of member and casual users") +
#                     scale_y_continuous(labels = scales::comma) +
#                     theme(legend.position = c(.10, .75), 
#                               legend.text = element_text(face = "bold"),
#                               plot.title = element_text(hjust = 0.5)) +
#                     labs(fill='User type') +
#                     xlab("month") + ylab("count") +
#                     scale_x_date(date_labels = "%b %y", date_breaks = "1 month")

plot_line_monthly_use <- ggplot(data = trip_data_clean_v2, aes(x = as.Date(started_at, "%Y-%m-%d %h:%m:%s", tz="UTC"), color=member_casual)) +
                    geom_line(stat='count', size = 2) +
                    ggtitle("Trips of member and casual users in the previous 12 months") +
                    scale_y_continuous(labels = scales::comma) +
                    theme(legend.position = c(.80, .75), 
                              legend.text = element_text(face = "bold"),
                              plot.title = element_text(hjust = 0.5)) +
                    scale_color_discrete(name="User type") +
                    xlab("month") + ylab("count") +
                    scale_x_date(date_labels = "%b %y", date_breaks = "1 month")

#Day of the week of casual versus members
plot_day_of_week_use <- ggplot(data = trip_data_clean_v2, aes(x = week_day, fill=member_casual)) +
                        geom_bar(position="dodge") +
                        ggtitle("Trips for each day of the week") +
                        scale_y_continuous(labels = scales::comma) +
                        theme(legend.position = c(.15, .95), 
                                  legend.text = element_text(face = "bold"),
                                  plot.title = element_text(hjust = 0.5)) +
                        labs(fill='User type') +
                        xlab("weekday") + ylab("count")

#Plot time of the day of casual versus members
plot_hourly_use <- ggplot(data = trip_data_clean_v2, aes(x = hour(started_at), fill=member_casual)) +
                   geom_bar(position="dodge") + 
                   ggtitle("Trips for each hour in the day") +
                   scale_y_continuous(labels = scales::comma) +
                   scale_x_continuous(labels=c(0:23), breaks = c(0:23)) +
                   theme(legend.position = c(.15, .75), 
                            legend.text = element_text(face = "bold"),
                            plot.title = element_text(hjust = 0.5)) +
                   labs(fill='User type') +
                   xlab("hour") + ylab("count")
  
```

## Density maps of bike use

```{r, echo=FALSE}
if(!requireNamespace("devtools")) install.packages("devtools")
devtools::install_github("dkahle/ggmap")
library("ggmap")

# Enter your google maps key and uncomment the 2 lines below
#key_google <- <enter secret key here>
#register_google(key=key_google, write = TRUE)

# Try a Density Map
map_center <- geocode("1800 N Halsted St, Chicago, IL 60614") # Center map
chicago <- get_map(map_center, zoom = 12)
# Chicago_Map_pickup_topright <- ggmap(chicago, extent = "device", legend = "topright")
Chicago_map_pickup_bottom <- ggmap(chicago, extent = "device", legend = c("bottom")) # have legend on bottom

trip_data_random <- trip_data_clean_v2[sample(nrow(trip_data_previous_12m), 500000), ] # randomly select a subset of data

# density overlay
overlay_start <- stat_density2d(
    data = trip_data_random,
    aes(x = start_lng, y = start_lat, fill = ..level.., alpha = ..level..),
    size = 2, bins = 10,
    geom = "polygon"
  ) 

#facet_Wrap_days <- facet_wrap(~wday(trip_data_random$started_at), nrow=3)

# Density map of all users
# Map_MemberCasual_Density <- Chicago_Map_pickup_topright + 
#                               overlay_start + 
#                               scale_fill_gradient(low = "#FFCCCB", high = "#cc0000") + 
#                               labs(fill="Density") + guides(alpha = F)  +
#                               facet_wrap(~member_casual, ncol=2) + 
#                               ggtitle("Casual Users Previous 12 months")
# 
# # Density map of weekday
# Map_Weekday_Density <- Chicago_map_pickup_bottomright + 
#                               overlay_start + 
#                               scale_fill_gradient(low = "#FFCCCB", high = "#cc0000") + 
#                               labs(fill="Density") + guides(alpha = F)  +
#                               facet_wrap(~week_day, ncol=4)

# Facet Density Map of Casual Users and Members for each day of the week
Map_Weekday_MemberCasual_Density <- Chicago_map_pickup_bottom + 
                              overlay_start + 
                              scale_fill_gradient(low = "#FFCCCB", high = "#cc0000") + 
                              labs(fill="Density") + guides(alpha = F)  +
                              facet_wrap(facets = c(~member_casual, ~week_day), ncol=7.) +
                              ggtitle("Density Map of Day of the Week Use") +
                              theme(plot.title = element_text(hjust = 0.5))

```

## Conclusion

### Summary
- Casual users outnumber annual members on the weekend
- Annual members outnumber casual users Monday through Friday.
- Annual member use spikes between 6:00 to 8:00 and 16:00 to 19:00. Traditional communing hours
- Annual members have highest density in the central business district
- Casual users have more spread out density with hot spots in recreational areas

### Recommendations
- Market to casual users that use the service similar to annual members
  - Remind casual users of annual membership benefits
  - Promote the savings when becoming an annual member
- Conduct user survey through the app and in-person
  - Identify how people are using the service (commute, recreation)
- Test surge pricing for casual users during commuting and positive weather

